Estimating duration in the presence of misclassification

### 2023-06-07 cis-imperfect-testing.tex
Write first half of chapter

@@

In \cref{perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
\todo[inline]{Check exactly what has been done and verify exactly what I can claim here}
In this chapter, I modify the previous model to include false negatives (\cref{imperf-test:sec:modelling}).
I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
Plausible estimates of the duration distribution are produced, however, they require assuming that the test sensitivity (the probability of testing positive given being detectable) is known; I discuss what would be required to relax this assumption in \cref{imperf-test:sec:discussion}.

@@

Comparing the simulated and CIS data suggested that this is due to the high frequency of single positive episodes in the CIS data.
A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed that in a related setting that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
Furthermore, simulations building in even a small number of excess single positive episodes led to problematic estimation even within the CIS setting.
\todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}

@@

There are two strands of evidence that support this hypothesis.
First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{ATACCC}; they can be directly observed within CIS sa intermittent negatives (see below).
Second, adjusting the simulation to include false negatives can reproduce both the CIS data more realisticly and the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})

@@

The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
This can be simulated by...

However, an implausibly low test sensitivity is required to reproduce the rate of single positive episodes seen in CIS.
\todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
More plausibly, the rate of false negatives varies over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{ATACCC}).
\todo[inline]{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

Examination of the CIS data also provides evidence for a declining test sensitivity.
We can bound the test sensitivity by...
\todo[inline]{Include the analysis of CIS data with bounding test sensitivity}

A simple declining test sensitivty might be:
\todo[inline]{Equation for linearly decreasing test sensitivity}
Simulated data using this test sensitivity is much more similar ot the data, while being more plausible than very low probabilities of false negatives.

@@

Specifically, we introduce a constant test sensitivity $p_\text{sens}$
such that:
\begin{align}
\prob(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
\end{align}

In practice, it is likely that the test sensitivity is a function of the
viral load of an individual (the concentration of viral material in the
part of the body being sampled). For this initial attempt at modelling
the sensitivity, we ignore this complication.

>>>>

\newcommand{\psens}{p_\text{sens}}

@@

In \cref{perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
\todo[inline]{Check exactly what has been done and verify exactly what I can claim here}
In this chapter, I modify the previous model to include false negatives (\cref{imperf-test:sec:modelling}).
I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study-results}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
Plausible estimates of the duration distribution are produced, however, they require assuming that the test sensitivity (the probability of testing positive given being detectable) is known; I discuss what would be required to relax this assumption in \cref{imperf-test:sec:discussion}.

@@

Comparing the simulated and CIS data suggested that this is due to the high frequency of single positive episodes in the CIS data.
A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed that in a related setting that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
\todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}

@@

There are two strands of evidence that support this hypothesis.
First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})

@@

The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
Denoting the test sensitivity by $\psens$, we replace the deterministic $y_i(t)$ with a random binary variable such that:
\begin{equation}
  \prob(y_i(t) = 1) \begin{cases}
      \psens &b_i \leq t \leq e_i \\
      0 &\text{otherwise}
  \end{cases} 
\end{equation}
with $y_i(t)$ still only defined for $t \in t_i$.

However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
\todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
More plausibly, the rate of false negatives varies over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{ATACCC}).
\todo{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

Examination of the CIS data also provides evidence for a declining test sensitivity.
We can approximately bound the test sensitivity by assuming that all intermittent negatives are false negatives, that the negative following the last positive in an episode could be either a true or false negative, and all other negatives are true negatives.
The point estimate for the test sensitivity is the true positives / (true positives + false negatives).
In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
While the bounds are broad, they clearly suggest a declining test sensitivity over the course of an episode.
\begin{figure}
  \todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
  \caption{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
\end{figure}

I propose the following model for test sensitivity compatible with \cref{imperf-test:fig:bounding-cis-sensitivity}, consisting of a linearly declining period then a constant period:
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
\end{equation}
where $t$ is the time in days since infection
Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
\todo{Figure comparing the variable test sensitivity to fixed and CIS data}

@@

The model is only extended to model a constant test sensitivity.
Further work is required to introduce the more realistic varying test sensitivity (discussed in \cref{imperf-test:sec:discussion}).
